Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-10-18 paper/type-system.tex
Edits.

Fig~\ref{fig:fb-syntax} describes the syntax of our formal language,
which we call \fbname (\FB). The language is seeded with \ObjZ and
\RgnZ classes. More classes can be defined using the \C{class}
keyword. Class types in \FB are region-annotated variants of class
types in FGJ (also called \emph{core types}). This correspondence is
reflected in the $T@\rgn$ syntax of the region type of an object,
whose core type is $T$, and which is contained in the region $\rgn$.
We let $\rgn$ range over static identifiers of regions in \FB. Note
that all \FB objects are boxed values, hence their \FB types are
always region-annotated. The only unboxed value in \FB is $\unitval$
of type \unitZ.

The expression language of FGJ is impoverished, lacking, among others,
an assignment expression. Note that it is possible to \emph{encode}
assignments to instance variables (object fields) in FGJ by creating a
new object with its fields set to new values. Unfortunately, such
encoding is not semantics preserving in case of \name, as the new
object could be allocated in a region different from the allocation
region of the original object. To avoid such complications, we extend
the expression language with assignments and local variable
declarations (via \C{let} expressions). Instead of distinguishing
between statements (e.g., assignments) and expressions (e.g., method
calls) in \FB, we consider statements as expressions of \unitZ type.
Statements that introduce lexical blocks in \name, such as
\C{letregion} and \C{open} are, are converted to their expression form
using the \C{in} keyword.

The expression language has been equipped with lambda abstraction
($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
expressions to define and apply anonymous functions.  Modulo the angle
braces containing $\rho$'s and $\phi$'s, whose presence will be
justified later, these expressions are uncurried variants of
corresponding expressions from simply typed lambda calculus. Note that
FGJ formalism does not include higher-order functions, hence the
syntactic class (\C{T}) of FGJ types needs to be extended with an
(uncurried) arrow type.

Informally, \FB class definitions are extensions of FGJ class
definitions with regions. A class definition in FGJ can be lifted to
its \FB definition by (a). lifting all core types in the body of the
class (i.e, fields, constructors and methods) to their
region-annotated versions, and (b). parameterizing the class over
region variables that occur free in the region-annotated body of the
class. Note that in \FB, regions, and their static identifiers
($\rgn$) are confined to the expression language. Consequently, the
region annotated types of class fields, constructors and methods do
not refer to identifiers of concrete regions, but rather their place
holders called \emph{region variables} ($\rho$). Informally, region
variables ($\rho$) are to region identifiers ($\rgn$) as type
variables ($a$) are to types ($T$).

Note that when a generic class definition in FGJ is lifted to a
region-polymorphic definition in \FB, it nonetheless remains generic
with respect to core types. We say that the class is now a
region-polymorphic generic class. Region parameterization of a class
in \FB is independent of its parameterization over core types; they
are not conflated to make the class parametric over region types. For
instance, following is the definition of region-polymorphic generic
\C{Pair} class (The symbol $\extends$ should be read \emph{extends}):
\begin{codejava}[mathescape=true]
class Pair<$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a 
                      \conj \rho_2 \outlives \rho^a$>
          <a $\extends$ Object, b $\extends$ Object> {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
    super(); 
    this.fst = fst; 
    this.snd = snd;
  }
  a@$\rho_1$ getFst() {
    return this.fst;
  }
}
\end{codejava}
Every class definition in \FB is necessarily polymorphic with respect
to the allocation region of its objects. As a rule, allocation region
parameter (denoted $\rho^a$) is the first region parameter of a class.
Besides the allocation region of its objects, \C{Pair} class is also
parametric over the regions its first and second elements are
allocated in. References between objects allocated in different
regions are only allowed if the referred object is guaranteed to
outlive the referring object. In case of \C{Pair} class, this means
that allocation regions ($\rho_1$ and $\rho_2$) of both objects that
make up the pair must outlive the allocation region ($\rho^a$) of the
\C{Pair} object. Such conditions over region parameters of a class
need to be recorded in its header as region constraints ($\phi$) in
order for the class to be judged well-formed by the type system
(Fig.~\ref{fig:fb-morewfrules}).

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers ($\rgn$), respectively. Region instantiation has to
satisfy the stated constraints over region parameters. For example,
the following code snippet appropriately instantiates region
parameters of the \C{Pair} class to construct a \C{Pair} of
{\ObjZ}{\!}s, each allocated in different regions:
\begin{codejava}
letregion $\rgn_0$ {
  Object<$\rgn_0$> fst = new Object<$\rgn_0$>();
  letregion $\rgn_1$ {
    Object<$\rgn_1$> snd = new Object<$\rgn_1$>();
    Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$> p = 
       new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>(fst,snd);
  }
}
\end{codejava}
Since $\rgn_0 \outlives \rgn_1 \conj \rgn_1 \outlives \rgn_1$, the
region constraints of the \C{Pair} class are satisfied. Observe that
the region type of pair object \C{p} conveys the fact that (a). it is
allocated in region $\rgn_1$, and (b). it holds references to
object(s) allocated in region $\rgn_0$. On the other hand, if we
choose to allocate the \C{fst} object also in $\rgn_1$, the region
type of \C{p} would be
\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which can be
written simply as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. In general, the
\C{@} notation in a region type of an object \C{x} highlights that
\C{x}, and all the objects reachable from \C{x} via references are
allocated in a single region. We say that \C{x} is \emph{contained} in
the region. The objects of $\ObjZ$ class, by the virtue of
containing no references to other objects, are always contained in
their allocation region. Hence the type $\ObjZ@\rgn$ is equivalent to
the type $\ObjZ\inang{\rgn}$.
%%An object \emph{allocated} in a region $\rgn$ contains its spine in
%%$\rgn$, but can refer to objects allocated in other regions.  On the
%%other hand,

>>>>

Fig~\ref{fig:fb-syntax} describes the syntax of our formal language,
which we call \fbname (\FB). The language is seeded with \ObjZ and
\RgnZ classes. More classes can be defined using the \C{class}
keyword. Class types in \FB are region-annotated variants of class
types in FGJ (also called \emph{core types}). This correspondence is
reflected in the $T@\rgn$ syntax of a region type, which is the
simplest form of a region type describing an object, of core type $T$
contained in a region $\rgn$. We let $\rgn$ range over static
identifiers of regions in \FB. Note that all \FB objects are boxed
values, hence their \FB types are always region-annotated. The only
unboxed value in \FB is $\unitval$ of type \unitZ.

The expression language of FGJ is impoverished, lacking, among others,
an assignment expression. While it is possible to encode
assignments to instance variables (object fields) in FGJ by creating a
new object with its fields set to new values, such encoding is not
semantics preserving in case of \name, as the new object could be
allocated in a region different from the allocation region of the
original object. To avoid such complications, we extend the expression
language with assignments and local variable declarations (via \C{let}
expressions). Instead of distinguishing between statements (e.g.,
assignments) and expressions (e.g., method calls) in \FB, we consider
statements as expressions of \unitZ type.  Statements that introduce
lexical blocks in \name, such as \C{letregion} and \C{open} are, are
converted to their expression form using the \C{in} keyword. We let
$\rgn$ range over static region identifiers introduced by these
expressions.

The expression language has been equipped with lambda abstraction
($\lambdaexp{...}{\xbar : \taubar}{e}$) and application ($e(\bar{e})$)
expressions to define and apply anonymous functions. Modulo the angle
braces, these expressions are uncurried variants of corresponding
expressions from System F. Note that FGJ formalism does not include
higher-order functions, hence the syntactic class (\C{T}) of FGJ types
needs to be extended with an (uncurried) arrow type.

Informally, \FB class definitions are extensions of FGJ class
definitions with regions. A class definition in FGJ can be lifted to
its \FB definition by (a). lifting all core types in the body of the
class (i.e, fields, constructors and methods) to their
region-annotated versions, (b). parameterizing the class over
region variables that occur free in the region-annotated body of the
class, and (c). recording any outlives constraints on the region
parameters of the class. The later two steps effectively make the
class region polymorphic (cf. \S\ref{sec:alloc-ctxt}: qualified region
polymorphism). Note that in \FB, regions, and their static identifiers
($\rgn$) are confined to the expression language. Consequently, the
region annotated types of class fields, constructors and methods do
not refer to identifiers of concrete regions, but rather their place
holders called \emph{region variables} ($\rho$). Informally, region
variables ($\rho$) are to region identifiers ($\rgn$) as type
variables ($a$) are to types ($T$).

Note that when a generic class definition in FGJ is lifted to a
region-polymorphic definition in \FB, it nonetheless remains generic
with respect to core types. We say that the class is now a
region-polymorphic generic class. Region parameterization of a class
in \FB is independent of its parameterization over core types; they
are not conflated to make the class parametric over region types. For
instance, following is the region-polymorphic definition of a generic
\C{Pair} class (The symbol $\extends$ should be read \emph{extends}):
\begin{codejava}[mathescape=true]
class Pair<$\rho^a, \rho_1, \rho_2 \,|\, \rho_1 \outlives \rho^a 
                      \conj \rho_2 \outlives \rho^a$>
          <a $\extends$ Object, b $\extends$ Object> {
  a@$\rho_1$ fst; 
  b@$\rho_2$ snd;
  Pair(a@$\rho_1$ fst, b@$\rho_2$ snd) {
    super(); 
    this.fst = fst; 
    this.snd = snd;
  }
  a@$\rho_1$ getFst() {
    return this.fst;
  }
}
\end{codejava}
When objects of a class are allocated in a region $\rgn$, it means
that the class's constructor is run with $\rgn$ as its allocation
context. Every class definition in \FB is necessarily polymorphic with
respect to the allocation region of its objects, i.e., the allocation
context of its constructor. We adopt a convention that requires the
allocation region parameter (denoted $\rho^a$) to be the first region
parameter of a class.  Besides the allocation region of its objects,
\C{Pair} class is also parametric over the regions its first and
second elements are allocated in. References between objects allocated
in different regions are only allowed if the referred object is
guaranteed to outlive the referring object. In case of \C{Pair} class,
this means that allocation regions ($\rho_1$ and $\rho_2$) of both
objects that make up the pair must outlive the allocation region
($\rho^a$) of the \C{Pair} object. Such conditions over region
parameters of a class need to be recorded in its header as region
constraints ($\phi$) in order for the class to be judged well-formed
by the type system (Fig.~\ref{fig:fb-morewfrules}).

To construct objects of the \C{Pair} class, its type and region
parameters need to be instantiated with core types ($T$) and concrete
region identifiers ($\rgn$), respectively. Region instantiation has to
satisfy the stated constraints over region parameters. For example,
the following code snippet appropriately instantiates region
parameters of the \C{Pair} class to construct a \C{Pair} of
{\ObjZ}{\!}s, each allocated in different regions:
\begin{codejava}
letregion $\rgn_0$ {
  Object<$\rgn_0$> fst = new Object<$\rgn_0$>();
  letregion $\rgn_1$ {
    Object<$\rgn_1$> snd = new Object<$\rgn_1$>();
    let p = new Pair<Object,Object><$\rgn_1$,$\rgn_0$,$\rgn_1$>
                  (fst,snd);
  }
}
\end{codejava}
Since $\rgn_0 \outlives \rgn_1 \conj \rgn_1 \outlives \rgn_1$, the
region constraints of the \C{Pair} class are satisfied, hence the
instantiation is valid. Observe that the region type of \C{p} conveys
the fact that (a). it is allocated in region $\rgn_1$, and (b). it
holds references to objects allocated in region $\rgn_0$ and $\rgn_1$.
In contrast, if we choose to allocate the \C{fst} object also in
$\rgn_1$, the region type of \C{p} would be
\C{Pair<\ObjZ,\ObjZ><$\rgn_1$,$\rgn_1$,$\rgn_1$>}, which can be
written simply as \C{Pair<\ObjZ,\ObjZ>@$\rgn_1$}. In general, the
\C{@} notation in a region type of an object \C{x} highlights that
\C{x}, and all the objects reachable from \C{x} via references are
allocated in a single region. We say that \C{x} is \emph{contained} in
the region. The $\ObjZ$ class in \FB contains no references to other
objects, hence its objects are always contained in their allocation
region. Their region type is $\ObjZ\inang{\rgn}$ (or equivalently,
$\ObjZ@\rgn$), for some region $\rgn$.
%%An object \emph{allocated} in a region $\rgn$ contains its spine in
%%$\rgn$, but can refer to objects allocated in other regions.  On the
%%other hand,

\FB's $\RgnZ$ objects, like $\ObjZ$ objects, have region type of form
$\RgnZ\inang{\rgn}$. However, $\rgn$ cannot be any region. Recall
that $\RgnZ$ objects have special semantics in \name - they act as
handlers to transferable regions. Constructing a new $\RgnZ$ object
entails the creation of a new transferable region, and it is in this
region that the new object is allocated in. It follows that $\rgn$ in
$\RgnZ\inang{\rgn}$ should be the static identifier of the new
transferable region. But, static identifiers for transferable
regions are introduced only when such regions are opened for
allocation via \C{open} expression. What, then, should be the region
type of a \C{new Region} expression?

\FB resolves this problem by existentially quantifying the allocation
region of $\RgnZ$ objects when they are created. In other words, the
type of \C{new Region} expression in \FB is
$\exists\rho.\RgnZ\inang{\rho}$. Existential quantification in the
type captures the fact that there now exists a transferable region
containing the newly constructed $\RgnZ$ object. Elimination of
existential quantification is facilitated by the \C{open} expression,
which opens the transferable region and assigns it an identifier
($\rgn$).  Within the scope of \C{open}, the transferable region is
identified with $\rgn$, allowing its handler to instantiate the
existentially bound region variable with $\rgn$, and assume the type
of $\RgnZ\inang{\rgn}$.

Existentially quantifying the allocation region in the type of $\RgnZ$
objects is also useful to assign appropriate types to data structures
of transferable regions. For example, a list of transferable regions
has the type $\C{List}\inang{??}$ in \FB.
